Light inextensible strings (thread) under tension in the Schwarzschild geometry

### 2021-02-28 tpe_edreview_rebuttal.txt

The referee is absolutely correct here and raises an important point.
When I said that the circumference was "trivially" $2\pi r$, what I
meant was that integrating equation (1) from $phi=0$ to $2\pi$ at
fixed $t,r,\theta$ was trivial.  The resulting proper string length is
a direct consequence of the *definition* of "r" [which is sometimes
called the circumferential radial coordinate].

This comment has made me realise that the $2\pi r$ proper string
length is actually the simplest proper length calculation possible in
the Schwarzschild metric, and I have rephrased accordingly, and I hope
it is more instructive.  I remember struggling with this myself as an
undergraduate (I wanted to know what the "real" radius was ...).
Einstein himself said "it is not so easy to free oneself from the idea
that coordinates must have an immediate metrical meaning", so I have
good company!

> 6.  Question.  There may be a typographical error in Eq. (2).
> Shouldn't the coefficient of the logarithm be 2M instead of 1/2?  As
> written the equation is not dimensionally consistent, since r and M
> are measured in the same units.  Compare to Taylor & Wheeler's
> Exploring Black Holes, p. 2-28, Eq. (27).
>

Factor of M now added...actually this caused me some considerable
confusion and uncertainty.  The equation presented in Taylor and
Wheeler has a factor of 2M but it should be M, for which I offer the
following sage transcript:

sage: var('r M r1 r2') 
(r, M, r1, r2)
sage: assume(r>0,r1>0,r2>0, r1>2*M, r2>2*M, r2-r1>0, M>0) 
sage: integrate(1/sqrt(1-2*M/r),r) 
M*log(sqrt(-2*M/r + 1) + 1) - M*log(sqrt(-2*M/r + 1) - 1) + r*sqrt(-2*M/r + 1)

In the same section Taylor and Wheeler discuss one-centimeter long
rulers laid end to end many times, a good metaphor for proper length.

### 2021-02-28 tpe_edreview_rebuttal.txt

Herewith my rebuttal to the review of "Light inextensible strings..."
The referee's comments are preceded by a chevron (">") and my response
is inline.  I have tinkered with linebreaks here and there in the
interests of clarity, and have replaced some mathematical notation in
the original review with latex equivalents.

Short story: I agree with all the comments made and have modified the
manuscript accordingly.  The document is stronger, more cohesive, more
accurate, and hopefully more instructive.  I submit that the extra
clarity and pedagogical value more than compensates for the slight
extra length.

I have changed the discussion section at the end very slightly
(emphasising the staticness of the metric considered) as a result of
reflecting on the totality of the comments too.

@@

> 
> 2.  Suggestion about the line that begins on p. 1 and continues to
> p. 2, about "lengthscales $L\leq T/(\mu g)$: Saying it this way may
> present a sticking point for the undergraduate audience towards whom
> this paper in intended.  The logic behind this inequality would show
> greater clarity if one says something like "the weight of the string
> does not exceed the tension, so that $L\mu g<T$".  From this the
> stated inequality about lengthscales follows easily.
>

This is a much better way of phrasing the argument and I have
rephrased as suggested.

> 3.  Suggestion.  Stating "$c^4/4GM$ for the strength of gravity at the
> event horizon" may leave an undergraduate audience baffled.  By
> starting from where they are, with $g=GM/r^2$, then setting r equal to
> the Schwarzschild radius, the stated result follows immediately,
> without mystery.
>

I have totally rephrased that part of the manuscript using this
suggestion and it is much clearer (I have also changed uppercase L for
lowercase l as per suggestion number 7).

@@

> 10. Suggestion.  Line below Eq. (5), I think the word "dashes" should
> be "primes".  This is a dash: --- and this is a prime: ', at least in
> usage with which I am familiar.

Fixed, notation now consistent

> 
> 11.  The paragraph below Eq. (6) is puzzling.  Here the path of the
> tight string is being compared to an "equatorial null geodesic," which
> I take to mean the path of a light beam moving along the equator.  If
> we are considering a light beam, then shouldn't a term with $t$ in it
> appear somewhere?  The puzzlement continues given the statement that
> "light, being sensitive to the time component of the metric" whereas
> the equations discussed in this paragraph say nothing (that I can see)
> about the time component of the metric.  The proper time interval
> would be zero, but the coordinate time interval $dt$ would, I think,
> be nonzero.  What am I missing?
>

The equation given as $u''=3u^2/2-u$ [here $u=2M/r$] gives the path of
a beam of light through Schwarzschild space in which $r$ is given as a
function of angle, $r=r(\phi)$.  The idea is that a stationary
observer has a laser pointer pointing in a particular direction,
switches it on and leaves it on for a long time.  We then record all
the points that the beam touches.  The path of the light is then
simply $r=r(\phi)$, no time coordinate involved, as Figures 4 and 6
show.  I have changed the citation to Rindler, who gives a detailed
derivation.

@@

> 12.  In the passage below Eq. (6), I am having trouble making the
> connection between $r''(\phi)=r-3M + 2(r')^2/r$ and the claim that the
> solution is $r=r_0\sec(\phi-\phi_0)$.  What trig identity am I
> missing?
>

There are two separate issues here.  Firstly, null geodesic paths in
the Schwarzschild metric obey $u''=3u^2/2-u$ where $u=2M/r$.  This is
equivalent to $r''(\phi) = r-3M + 2(r')^2/r$.  The second issue is
straight lines in Euclidean flat space (in flat space the expression
"straight line" is unambiguous).  WLOG $\phi_0=0$ and if
$r=r_0\sec\phi$ then $r'=-r_0\sin\phi\cdot\sec^2\phi$ and
$r''=r_0(\sec\phi + 2\sin^2\phi\sec^3\phi$, from which
$r''=r+2(r')^2/r$ follows directly.

### 2021-03-11 tpe_edreview2_rebuttal.tex

\documentclass[12pt]{article}
\usepackage{xcolor}
\begin{document}

I appreciate the author responding to each of my previous questions
and suggestions. For the most part I am satisfied with the author’s
responses and subsequent modifications to the manuscript. While it is
not my purpose to annoy the author or delay an editorial decision on
publication, a small number of comments about minor points remain:

1. The discussion about circumference and $2\pi r$ may be, to the
undergraduate reader to whom the Schwarzschild metric is still quite
new, a source of confusion regarding the logic behind the metric. In
the revised manuscript we read of ``constant $r$ whose proper length
is given by integration," where the integral is

\[
\int_{0}^{2\pi} r\,d\phi = 2\pi r
\]

While the integration is certainly true, it this context it appears to
express a tautology because the $r$- coordinate is defined as the
circumference of a shell (centered on the origin) divided by $2\pi$.
The passage just quoted makes it sound (at least to me) like
``circumference = $2\pi r$ is a derived result, when in fact it the
logic goes the other way---the measured circumference provides the
definition of $r$ (see Taylor \& Wheeler, Black Holes, page 2-7, which
in T\&W pagination means page 7 in Ch. 2; T\&W is a popular
entry-level textbook for undergraduates interested in general
relativity).

2. The author's explanation registers with more clarity now in the
passage about string being ``arbitrarily" large.  At first reading
this sounds like the author says that the length of a tight string
fixed between two points may be arbitrary, notwithstanding the
distinction between ``arbitrary" and ``arbitrarily."  Here is what I
now understand the author to be saying about the value of l in flat
and in Schwarzschild spacetimes:

\[
\ell_\mathrm{flat} = 2Mw
\]

and, asymptotically,

\[
\ell_\mathrm{Sch}\sim 2Mw + \log w
\]

and thus their difference is

\[
\ell_\mathrm{Sch} - \ell_\mathrm{flat}\sim M\log w
\]

which goes to infinity logarithmically; in other words, for an
infinite string the Schwarzschild metric is a ``worse infinity" which
diverges faster than what occurs in flat spacetime.  So far, so good.
I guess what bothers me is the statement that ``even far from [emphasis
  added] the event horizon, radial strings are arbitrarily longer than
flat space would imply...".  So long as $w$ is large (however far from
the event horizon) but finite, the lengths in both metrics are precise
numbers, not arbitrary.  Only when $w$ goes to infinity does the
length in Schwarzschild spacetime become infinitely longer than the
length in flat spacetime, although though both are infinite.  I
suppose this passage would sit better with this reader by merely
stating that the asymptotic difference between $\ell_\mathrm{Sch}$ and
$\ell_\mathrm{flat}$ diverges logarithmically.  But perhaps this is
only a matter of taste.

3. I still must disagree with the author about the coefficient of the
logarithm term in Eq. (2).  The original manuscript put this
coefficient at $1/2$; I argued that it is $2M$ because of the
corresponding equation in T\&W (their Eq. [27] on p. 2-28).  I also
find disagreement with Eq. (2) on a sign in the denominator of the log
term.  Standing ready to be corrected, please allow me to show why
this reader thinks these discrepancies exist.  We are to evaluate

\[
\ell=\int_{r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}} =
\int_{r_1}^{r_2}\frac{\sqrt{r}\,dr}{\sqrt{r-2M}},
\]

Let $r=z^2$ so that $dr=2z\,dz$, and let $2M=a^2$.  Then

\[
\ell=2\int_{r_1}^{r_2} \frac{z^2\,dz}{\sqrt{z^2-a^2}}.
\]

 Using the integral table in the CRC handbook, integral No. 171,

\[
\int\frac{dx}{\sqrt{x^2\pm a^2}}=\frac{x}{2} \sqrt{z^2\pm a^2}
\mp \frac{a^2}{2}\log\left| z+\sqrt{x^2\pm a^2}\right|
\]

 and with the minus sign under the radical in the integrand, I obtain

\[
\ell = 2\left[
  \frac{z}{2}\sqrt{z^2-a^2} + \frac{a^2}{2}\log\left|z+\sqrt{z^2-a^2}\right|
  \right]_{z_1}^{z_2}
\]

\[
=
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} + 2M\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}\right|
\]

In contrast, Eq (2) reads (points of disagreement in red):

\[
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} 
+\color{red}M
\color{black}\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}
  \color{red}-
  \color{black}\sqrt{r_1-2M}}\right|
\]

Other than these minor points, I have no further concerns about the
paper. I find the project very interesting, offering much to think
about. Having worked with undergraduate physics majors who wish to
learn some general relativity, once such students are made familiar
with the Schwarzschild metric, this exercise with a tight string
offers a useful tool. I leave it to the editors and authors to decide
what to do with these my latest suggestions regarding the revised
manuscript. I thank the author for taking my previous suggestions and
concerns seriously.

\end{document}

### 2021-03-12 tpe_edreview2_rebuttal.tex

1. The discussion about circumference and $2\pi r$ may be, to the
undergraduate reader to whom the Schwarzschild metric is still quite
new, a source of confusion regarding the logic behind the metric. In
the revised manuscript we read of ``constant $r$ whose proper length
is given by integration," where the integral is

\[
\int_{0}^{2\pi} r\,d\phi = 2\pi r
\]

While the integration is certainly true, it this context it appears to
express a tautology because the $r$- coordinate is defined as the
circumference of a shell (centered on the origin) divided by $2\pi$.
The passage just quoted makes it sound (at least to me) like
``circumference = $2\pi r$ is a derived result, when in fact it the
logic goes the other way---the measured circumference provides the
definition of $r$ (see Taylor \& Wheeler, Black Holes, page 2-7, which
in T\&W pagination means page 7 in Ch. 2; T\&W is a popular
entry-level textbook for undergraduates interested in general
relativity).

\textcolor{blue}{This is a perfectly reasonable comment and 
  I have changed the manuscript.\\
{\bf Old text:}
 ...(this is why $r$ is sometimes called ``circumferential radius'').\\
{\bf New text:} ...(this is the motivation for the definition of $r$;
indeed, it is  sometimes called ``circumferential radius'' for this
reason).
}

@@

and thus their difference is

\[
\ell_\mathrm{Sch} - \ell_\mathrm{flat}\sim M\log w
\]

which goes to infinity logarithmically; in other words, for an
infinite string the Schwarzschild metric is a ``worse infinity" which
diverges faster than what occurs in flat spacetime.  So far, so good.
I guess what bothers me is the statement that ``even far from [emphasis
  added] the event horizon, radial strings are arbitrarily longer than
flat space would imply...".  So long as $w$ is large (however far from
the event horizon) but finite, the lengths in both metrics are precise
numbers, not arbitrary.  Only when $w$ goes to infinity does the
length in Schwarzschild spacetime become infinitely longer than the
length in flat spacetime, although though both are infinite.  I
suppose this passage would sit better with this reader by merely
stating that the asymptotic difference between $\ell_\mathrm{Sch}$ and
$\ell_\mathrm{flat}$ diverges logarithmically.  But perhaps this is
only a matter of taste.

\textcolor{blue}{ This is a subtle point which took me some
  considerable time to grok.
}

@@

\[
\ell=2\int_{r_1}^{r_2} \frac{z^2\,dz}{\sqrt{z^2-a^2}}.
\]

 Using the integral table in the CRC handbook, integral No. 171,

\[
\int\frac{dx}{\sqrt{x^2\pm a^2}}=\frac{x}{2} \sqrt{z^2\pm a^2}
\mp \frac{a^2}{2}\log\left| z+\sqrt{x^2\pm a^2}\right|
\]

{\color{blue} The RHS of the equation above is quoted correctly, but
  the left hand side should be $\int\frac{x^2\,dx}{\sqrt{x^2\pm
      a^2}}$, not $\int\frac{dx}{\sqrt{x^2\pm a^2}}$ as quoted [this
    is CRC equation 164 in my edition].  }

@@

\[
=
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} + 2M\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}+\sqrt{r_1-2M}}\right|
\]

In contrast, Eq (2) reads (points of disagreement in red):

\[
\sqrt{r_2(r_2-2M)}-\sqrt{r_1(r_1-2M)} 
+\color{red}M
\color{black}\log\left|
\frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_1}
  \color{red}-
  \color{black}\sqrt{r_1-2M}}\right|
\]

{\color{blue} The reviewer is correct here but I would observe that
  the expression above (with or without the red alternative) is not
  equation 2 of the manuscript.  The reviewer's expression actually
  matches the RHS of Equation 2 in the manuscript which is

  \[
  \left.
  \sqrt{r(r-2M)} +M\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}
  \]

  \[
  = 
  \sqrt{r_2(r_2-2M)}-  \sqrt{r_1(r_1-2M)}
 +M\log\left(
 \frac{\sqrt{r_2}+\sqrt{r_2-2M}}{\sqrt{r_2}-\sqrt{r_2-2M}}
 \frac{\sqrt{r_1}+\sqrt{r_1-2M}}{\sqrt{r_1}-\sqrt{r_1-2M}}
 \right)
  \]

The match is not obvious: one needs to manipulate logarithms of
negative numbers and keep track of two cut lines on the complex plane
which manifest as (imaginary) constants of integration.  The apparent
factor of 2 resolves itself because multiplying numerator and
denominator by the appropriate conjugate surd results in the {\em
  square} of the expression given by the referee.

Arguably the referee's formulation is neater and more compact than
that appearing in the manuscript but I would support my own version on
the grounds that it is closer to mathematica's output (and indeed is
much more amenable to asymptotic analysis).

{\bf Summary: the reviewer's algebra was correct but our expressions
  in fact agree.}
}

### 2021-03-12 string_physics_educator_revision4.tex
work on circumferential radius

The simplest configuration for a string would be circular at
constant~$r$ whose proper length is given by integration of
equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
(this is why $r$ is sometimes called ``circumferential radius'').
Note carefully that $r$ is not the distance {\em to} anything, so this
result---which operates in Schwarzschild geometry and not plane
Euclidean geometry---is less obvious than it might appear.
Considering next a purely radial string we may follow Taylor and
Wheeler \cite{taylor2000} and calculate the proper length
$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:

>>>>

The simplest configuration for a string would be circular at
constant~$r$ whose proper length is given by integration of
equation~\ref{schwarzschild} as~$\int_{\phi=0}^{2\pi}r\,d\phi=2\pi r$
(this is the motivation for the definition of $r$; indeed, it is
sometimes called ``circumferential radius'' for this reason).  Note
carefully that $r$ is not the distance {\em to} anything, so this
result---which operates in Schwarzschild geometry and not plane
Euclidean geometry---is less obvious than it might appear.
Considering next a purely radial string we may follow Taylor and
Wheeler \cite{taylor2000} and calculate the proper length
$\ell\left(r_1,r_2\right)$ of a string stretching from~$r=r_1\geqslant
2M$ to~$r_2\geqslant r_1$ at constant $\theta,\phi$:
